### 2020-06-23 text/article_vedrovice.Rmd
results - spatial patterning

## Spatial pattern

Spatial arrangement of prehistoric burial grounds is often examined to find
any internal structure, arrangement or possible grouping.
We approach the task as a point pattern analysis problem given that
a burial (event) is simplified as a point location in geographic space.
A polygon of the excavation extent is used as a study area (a window).
Using a window closely approximating locations of the burials was also considered,
e. g. buffer zone or convex/concave hull but simplified excavation polygon is
deemed the most useful because it preserves the *natural* borders and empty areas
in the burial ground.

Overall structure of the burial ground is examined using both density based approach
and distance based methods [@osullivan2010, 121-154].
It is argued that density based approaches, for instance quadrat tests,
characterize the point pattern in terms of its *first-order properties* [@osullivan2010, 126].
The assumption underlying the *first-order effects* is that any point location is
independent on other point locations and different parameters influence the location
[@nakoinz2016, 130-135; @osullivan2010, 107].
On the other hand the methods based on distances between nearest neighbours and triples
characterize the point pattern based on *second- and third-order effects*.
In this case it is presumed that the point location depends on other
point locations.
<!-- [@nakoinz2016, 135-144; @osullivan2010, 130-139] -->

We assume that in case of a burial ground *first-order properties* are
of little to no importance at all.
In spatial analysis of settlement patterns individual locations of
settlements (points) can depend on a number of parameters, for instance
elevation, soil type etc.
<!-- It is possible that the position of burials was dependent on parameters that were not
captured by the excavation, for example location of trees but that is impossible to test.-->
In case of a grave placement in a burial ground we presume that the choice of
a location is influenced predominantly by other (previous) grave locations.
The *second-order properties*, and possibly *third-order properties* [@nakoinz2016, 130, 144]
are thus considered major factors in locating of the graves.

@@

Given statistics for independent random process are estimated using Monte Carlo
simulation with 99 iterations.
The most extreme results of simulations are discarded as outliers and the
second most extreme results are used to construct envelopes encompassing the
expected range for a given function [@baddeley2016 384-403]

@@

## Spatial pattern

```{r spat-gl, fig.cap = "G and L functions"}
knitr::include_graphics(here("plots", "pointprocess_fun.pdf"))
```

>>>>

## Spatial patterning

Spatial patterning in prehistoric burial grounds is often examined to find
any internal structure, arrangement or possible grouping.
We approach the task as a point pattern analysis problem given that
a burial (event) is simplified as a point location in a two dimensional space.
A polygon of the excavation extent is used as a study area (a window).
Using a window closely approximating locations of the burials was also considered,
e. g. buffer zone or convex/concave hull but simplified excavation polygon is
deemed the most useful because it preserves the *natural* borders and empty areas
in the burial ground.

Overall structure of the burial ground is examined using both density based approach
and distance based methods [@osullivan2010, 121-154].
It is argued that density based approaches, for instance quadrat tests,
characterize the point pattern in terms of its *first-order properties* [@osullivan2010, 126].
The assumption underlying the *first-order effects* is that any point location is
independent on other point locations and different parameters influence the location.
On the other hand the methods based on distances between neighbours and triples of neighbors
characterize the point pattern based on *second- and third-order effects*
[@nakoinz2016, 130-135; @osullivan2010, 107].
<!-- [@nakoinz2016, 135-144; @osullivan2010, 130-139] -->

We presume that certain burial location is determined by other burial locations
and is not quite dependent on on other parameters.
In other words *first-order properties* are of little to no importance at all in
our point of view.
In spatial analysis of settlement patterns individual locations of
settlements (points) can depend on a number of parameters, for instance
elevation, soil type etc.
<!-- It is possible that the position of burials was dependent on parameters that were not
captured by the excavation, for example location of trees but that is impossible to test.-->
In case of a grave placement in a burial ground we presume that the choice of
a location is influenced predominantly by other (previous) grave locations.
The *second-order properties*, and possibly *third-order properties* [@nakoinz2016, 130, 144]
are thus considered major factors in locating of the graves.

@@

Given statistics for independent random process are estimated using Monte Carlo
simulation with 99 iterations.
The most extreme results of simulations are discarded as outliers and the
second most extreme results are used to construct envelopes encompassing the
expected range for a given function [@baddeley2016 384-403].

To explore whether neighboring individuals are more often of the same sex we
construct a Gabriel graph [@gabriel1969] with burial locations as nodes (Fig. \@ref(fig:gabriel)).
For each node, mean sex of the neighbors is determined and recorded.
To check how common the observed value is, we perform a permutation test
where sex of burials is randomly assigned while preserving the overall
proportion of given categories and the neighborhood struture determined by the
Gabriel graph.
The observed values of neighbors sex categories are compared against distributions
derived from the simulated data.

```{r gabriel, fig.cap = "Gabriel graph"}
knitr::include_graphics(here("plots", "plan_vedrovice_gabriel.pdf"))
```

@@

## Spatial patterning

The quadrat test suggests both random and clustered processes with p values
under the 0.05 alpha value but the $\chi^2$ approximation is inaccurate due to small
expected counts (for the details see supplementary information).
Fig. \@ref(fig:spat-funs) shows that the curves of $G$ and $F$ functions do not
deviate too much from the curves expected under independent random process nor
do they traverse the boundaries of the envelopes.
This tells us that both the cummulative nearest-neighbor distance (*G*) and
distance to any random location in the space of the window (*F*) are well
approximated by random processes and do not show evidence for spatial clustering.
On the other hand, both *K* and *T* curves have a steep rise and exceed the
envelope boundaries.
These functions thus show evidence for clustering process in the observed point
pattern.

```{r spat-funs, fig.cap = "Functions *G*, *F*, *K* and *T*. *Dashed line* -- theoretical curve under CSR, *solid line* -- empirical curve, *gray area* -- envelope."}
knitr::include_graphics(here("plots", "pointprocess_fun.pdf"))
```

```{r neigh-sex, fig.cap = "Mean neighbors in a given sex category. *Gray area* -- density estimation for the simulated data, *vertical line* -- observed value."}
knitr::include_graphics(here("plots", "prob_neighbour_sex.pdf"))
```
